%% LaTeX2e class for student theses
%% sections/content.tex
%% 
%% Karlsruhe Institute of Technology
%% Institute of Information Security and Dependability
%% Software Design and Quality (SDQ)
%%
%% Dr.-Ing. Erik Burger
%% burger@kit.edu
%%
%% Version 1.5, 2024-02-12

\chapter{The QEMU API}
As the goal of this paper is to extract data without having to change code of QEMU,
it is neccessary to take a look at the data QEMU offers by itself.
QEMU offers 3 kinds of semi-external APIs: The QEMU Human Monitor (QHM),
the QEMU Machine Protocol (QMP) and so-called "Internal QEMU APIs"\cite{internal}.

As the name suggests, the internal QEMU APIs are not useful for external use by themselves.
Instead, they can be used when modifying QEMU-code and then extracting data by the means of GDB or directly by the QEMU process.
As the goal is to use any standard QEMU implementation, these internal APIs will not be regarded further in this paper.

In the two following sections, both QMP and QHM are further explored. Both the basic functionality,
as well as the commands used for this paper get explained in detail.
Also the differences of both protocols, especially in regards to their capabilities,
get discussed.

\section{QEMU Machine Protocol (QMP)} \label{sec:QMP}
"The QEMU Machine Protocol (QMP) is a JSON-based protocol which allows applications to control a QEMU instance"\cite{qmp-description}, as described by the QEMU-Wiki itself.
The protocol was merged into the main branch of QEMU in version 0.13 in 2010, however to this day its status remains as "experimental"\cite{qmp-merge}.
The goal was, and still is, to provide the same data QHM provides, but in a format more accessible for machines.
However, this process is far from complete. To give an example, the patch to add the ability of reading Register contents was only introduced in 2021,
but to this day it wasn't merged into the mainline branch.
Also, this patch doesn't actually provide a valid JSON-object, but just passes the return of the QHM-command "info registers" into the "result"-string\cite{qmp-registers-patch}.

This makes QMP the preferred protocol, as the data is in an easily parseable format, which makes extraction very easy.
Also the returned data is generally consistent across architectures, and the documentation is very detailed.
But, as mentioned, QMP is the less powerful protocol when used by itself.
This however directly leads to the most important command\cite{qmp-commands}:

\paragraph{human-monitor-command}
This command is perhaps the most important one in all of this, as it allows to execute QHM-commands in a QMP-environment.
The QEMU reference explicitly marks this command as a "stop-gap" until all QHM-commands have their counterpart in QMP,
but, as mentioned before, that currently isn't necessary a priority.
Still, it's currently the only way to access certain data like register contents, even if in an ugly fashion.
Further details on how this command is used are described later on in \autoref{sec:QHM}.

\paragraph{qmp_capabilities}
This command must be sent at the beginning of any QMP session to activate the session.
On an unmodified QEMU instance, it returns an empty JSON-object.
The command is intended to show the client additional features of a modified QEMU-instance,
however in this paper this command serves no purpose except unlocking QMP.

\paragraph{query-cpus-fast}
Gives information about all available CPUs, namely their architecture,
their count, and the ID of the host thread simulating the CPU.
The commmand is a replacement for \emph{query-cpus}, which was removed and returned slightly different data.
The main purpose for this paper is to get the CPU count, as registers need to be queried individually for each CPU.

\paragraph{query-block}
Queries for all kinds of drives. Gives information about attached storage mediums and where their content is stored.
This of course gets used to back up the neccessary files later on.
Also shows empty devices like optical drives which currently don't contain any media.